---
title: "F.M.L. Assignment 2"
author: "Christine Donovan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Assignment 2: k-NN Classification for Universal Bank

This analysis uses k-nearest neighbors (k-NN) to predict whether a Universal Bank customer will accept a personal loan offer, based on customer demographic and account data.

# Load libraries

```{r}
library(class)
library(caret)
library(ggplot2)
library(tidyr)
```

# Load the raw data UniversalBank.csv file

```{r}
bank<-read.csv("UniversalBank.csv")
```

# Convert Education to a factor

```{r}
bank$Education <- factor(bank$Education)
#Build and apply dummy-variable recipie
dummy_blueprint <- dummyVars(~ Education, data = bank)
Education_dummies <- predict(dummy_blueprint, newdata = bank)
```

```{r}
head(Education_dummies)
```

# Bank and Education_dummies (with the three new ones) using cbind to combine into a single dataframe.

```{r}
bank_full<-cbind(bank, Education_dummies)
```

# Remove columns: Education, ID and Zip Code

```{r}
which(names(bank_full) %in% c("ID", "ZIP.Code", "Education"))
```

```{r}
bank_full<- bank_full[ , -c(1, 5, 8)]
```

# Partioning the data training (60%) and the validation set(40% )

```{r}
set.seed(123)
train_index <- sample(1:nrow(bank_full), 0.60 * nrow(bank_full))
length(train_index)
```

```{r}
train_data <- bank_full[train_index, ]
nrow(train_data)
```

# Partioning the validation set (40%) All rows except the train_index

```{r}
validation_data<-bank_full[-train_index, ]
nrow(validation_data)
```

# Separate predictors from target variables before normalizing

```{r}
train_data_predictors <- train_data[ , -which(names(train_data) %in% c("Personal.Loan"))]
ncol(train_data_predictors)
```

# Build and apply Normalization recipe

```{r}
norm_recipie <- preProcess(train_data_predictors, method = c("range"))
train_predictors <- predict(norm_recipie, newdata = train_data_predictors)
validation_predictors <- predict(norm_recipie, newdata = validation_data[ , -which(names(validation_data) %in% c("Personal.Loan"))])
summary(train_predictors)
```

# Build and separate Label vectors

```{r}
train_labels <- factor(train_data$Personal.Loan)
validation_labels <- factor(validation_data$Personal.Loan)
```

```{r}
summary(train_predictors)
```

```{r}
train_data_predictors <- train_data[ , -which(names(train_data) %in% c("Personal.Loan"))]
ncol(train_data_predictors)
```

# Build data frame for new customer and normalize it

```{r}
new_customer_q1 <- data.frame(
  Age = 40,
  Experience = 10,
  Income = 84,
  Family = 2,
  CCAvg = 2,
  Mortgage = 0,
  Securities.Account = 0,
  CD.Account = 0,
  Online = 1,
  CreditCard = 1,
  Education.1 = 0,
  Education.2 = 1,
  Education.3 = 0
)
```

```{r}
new_customer_q1
```

# Normalize new customer

```{r}
new_customer_norm <- predict(norm_recipie, newdata = new_customer_q1)
new_customer_norm
```

# Problem 1: Run k-NN with k=1. The classification result (0, not accepted)

```{r}
knn_result_q1 <- knn(train = train_predictors, test = new_customer_norm, cl = train_labels, k = 1)
knn_result_q1
```

# Problem 2. Step 1: Rebuild a combined training datafame (normalized predictors + labels) & run check

```{r}
train_knn_data <- data.frame(train_predictors, Personal.Loan = train_labels)
str(train_knn_data)
```

# Step 2. Set up train() to search for the best K. 
```{r}
Personal.Loan ~. 
knn_tune <- train(Personal.Loan ~ ., data = train_knn_data, method = "knn", tuneGrid = expand.grid(k = 1:20))
knn_tune
  
```
# Ggplot Observation
```{r}
ggplot(knn_tune$results, aes(x = k, y = Accuracy)) +
  geom_line() +
  geom_point() +
  geom_vline(xintercept = 5, linetype = "dashed", color = "red") +
  labs(title = "k-NN Accuracy vs. k (Bootstrap Resampling)",
       x = "k (number of neighbors)",
       y = "Accuracy") +
  theme_minimal()
```

# Problem 2 Answer: 

Although train() selected k=1 as having the highest bootstrapped accuracy, k=1 is susceptible to overfitting. Accuracy remains high and relatively stable through k=6-7, after which it begins a more consistent decline. Choosing k=6 sacrifices a small amount of accuracy (95.2% → 94.6%) in exchange for a model less sensitive to individual data points.  To avoid a possible tie, testing k=5 directly against the validation and test sets, it outperformed k=6 on every metric, which supports settling on k=5 as the final choice.

# Validation Set Performance: k=6 vs k=5
```{r}
k_values <- c(1, 5, 6)
results_list <- list()

for (k_val in k_values) {
  pred <- knn(train = train_predictors, test = validation_predictors, cl = train_labels, k = k_val)
  cm <- confusionMatrix(pred, validation_labels, positive = "1")
  results_list[[as.character(k_val)]] <- c(
    Accuracy = cm$overall["Accuracy"],
    Sensitivity = cm$byClass["Sensitivity"],
    Specificity = cm$byClass["Specificity"],
    Kappa = cm$overall["Kappa"]
  )
}

comparison_df <- do.call(rbind, results_list)
comparison_df <- data.frame(k = k_values, comparison_df)
comparison_df
```

# Comparison Across K Values
```{r}
comparison_long <- pivot_longer(comparison_df, 
                                  cols = c(Accuracy.Accuracy, Sensitivity.Sensitivity, 
                                           Specificity.Specificity, Kappa.Kappa),
                                  names_to = "Metric", values_to = "Value")

ggplot(comparison_long, aes(x = factor(k), y = Value, fill = Metric)) +
  geom_col(position = "dodge") +
  labs(title = "Model Performance Comparison Across k Values",
       x = "k (number of neighbors)",
       y = "Score") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

# Observation

Although k=1 is theoretically the most susceptible to overfitting, both the bootstrap resampling in the initial tuning step and a direct test on the held-out validation set show k=1 outperforming k=5 and k=6 across every metric, including accuracy (96.2%), sensitivity (70.3%), and Kappa (0.768). Given the size of this training set (3,000 observations), this suggests k=1 may generalize well here rather than overfitting to noise. However, because k=1's performance can be more sensitive to the particular train/validation split chosen, I've decided to stay with a more cautious choice k=5. For the remainder of this analysis, k=5 is used as the working choice, balancing performance with some margin against the instability k=1 can exhibit on unseen data


# Problem 3 Part 1

```{r}
knn_pred_validation <- knn(train = train_predictors, test = validation_predictors, cl = train_labels, k = 5)
length(knn_pred_validation)
```

# Problem 3 Confusion Matrix

```{r}
confusionMatrix(knn_pred_validation, validation_labels, positive = "1")
```
The k=5 model achieves 95.0% overall accuracy, with sensitivity (the ability to correctly identify actual loan-acceptors) at 52.5%, meaning the model still misses just under half of customers who would accept a loan offer, though this is an improvement over k=6 (49.0%). Specificity (99.78%) remains excellent on the non-acceptor class. Kappa (0.6549) also improved over k=6 (0.6162), indicating the k=5 model adds more value beyond chance once the class imbalance (Prevalence = 10.1%) is accounted for. Overall, k=5 outperformed k=6 on every metric in this comparison, not just sensitivity, suggesting it's a better-fitting neighborhood size for this dataset while also avoiding the tie-vote risk that comes with an even k.

# Problem 4

```{r}
knn_result_q4 <- knn(train = train_predictors, test = new_customer_norm, cl = train_labels, k = 5)
knn_result_q4
```

Using k=5, this customer is classified as 0 (loan not accepted). The same results were obtained under k=1 and k=6. Even though k values disagree on some customers overall (the k=5 validation confusion matrix shows 96 false negatives and 4 false positives, compared to 103/7 under k=6), all three choices of k agree on this particular customer's classification.

# Problem 5

```{r}
set.seed(456)
train_index2 <- sample(1:nrow(bank_full), 0.50 * nrow(bank_full))
train_data2 <- bank_full[train_index2, ]
nrow(train_data2)
```

```{r}
remaining_data <- bank_full[-train_index2, ]
nrow(remaining_data)
```

```{r}
set.seed(456)
validation_index <- sample(1:nrow(remaining_data), 0.60 * nrow(remaining_data))
validation_data2 <- remaining_data[validation_index, ]
nrow(validation_data2)
```

```{r}
test_data2 <- remaining_data[-validation_index, ]
nrow(test_data2)
```

# Normalization Step 1: Separate predictor from target for all three

```{r}
train_predictors2 <- train_data2[ , -which(names(train_data2) %in% c("Personal.Loan"))]
validation_predictors2 <- validation_data2[ , -which(names(validation_data2) %in% c("Personal.Loan"))]
test_predictors2 <- test_data2[ , -which(names(test_data2) %in% c("Personal.Loan"))]
```

# Step 2: Build the recipe from train_predictors2 only, apply to all three

```{r}
norm_recipie2 <- preProcess(train_predictors2, method = c("range"))
train_predictors2 <- predict(norm_recipie2, newdata = train_predictors2)
validation_predictors2 <- predict(norm_recipie2, newdata = validation_predictors2)
test_predictors2 <- predict(norm_recipie2, newdata = test_predictors2)
```

# Step 3: Build the labels for all three

```{r}
train_labels2 <- factor(train_data2$Personal.Loan)
validation_labels2 <- factor(validation_data2$Personal.Loan)
test_labels2 <- factor(test_data2$Personal.Loan)
```

```{r}
ncol(train_predictors2)
```

# Run k-NN predictions for each set

```{r}
knn_pred_train2 <- knn(train = train_predictors2, test = train_predictors2, cl = train_labels2, k = 5)
knn_pred_validation2 <- knn(train = train_predictors2, test = validation_predictors2, cl = train_labels2, k = 5)
knn_pred_test2 <- knn(train = train_predictors2, test = test_predictors2, cl = train_labels2, k = 5)
```

# Build the three confusion matrices

```{r}

print("Training Set Confusion Matrix")
confusionMatrix(knn_pred_train2, train_labels2, positive = "1")

print("Validation Set Confusion Matrix")
confusionMatrix(knn_pred_validation2, validation_labels2, positive = "1")

print("Test Set Confusion Matrix")
confusionMatrix(knn_pred_test2, test_labels2, positive = "1")
```

# Conclusion
Using k=5, the training set again shows the highest accuracy (96.4%) and sensitivity (63.9%), since the model is evaluated on data it was built from some of its own points can act as their own nearest neighbors, inflating apparent performance. Validation (94.33%) and test (95.5%) show lower accuracy and notably lower sensitivity (44.97% and 53.33% respectively), reflecting more realistic performance on unseen data. Specificity remains consistently high (99.7–99.9%) across all three sets, reflecting that the model handles the majority non-acceptor class easily but struggles more with the minority acceptor class it's being evaluated on here. Compared to the earlier k=6 results, k=5 improved sensitivity in all three sets (training: 60%→63.9%, validation: 42%→44.97%, test: 49%→53.33%), supporting the choice of k=5 as a modest but genuine improvement, while also eliminating the possibility of tied votes that an even k like 6 could produce.

